Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
Some wordsmithing and reorg of the MRF intro

\remark{Dropped the section on ``our background,'' but the information
  should reappear when and where we discuss barriers to entry in more
  detail. ---NR}

@@

%% break the viterbi types up and insert into the Plan7 explanation.
%% go over this with NR; I can't make it flow.
% 
% \begin{figure}
% \lstinputlisting[label=viterbi_types,caption=Viterbi Types]{viterbi_types.hs}
% \end{figure}
%

@@

\emph{[This might be a place for comparison with standard
implementations of Viterbi.]}

@@

Our implementations of hidden Markov models and Viterbi's algorithm
simply duplicate 
functionality that is already implemented in the C++~package
HMMER.
These implementations form is just one component of our new tool, MRFy.

@@

\emph{[Next up: how does this bonding phenomenon constrain sequences
of homologous proteins?]}

In particular, these approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
Because of their pairwise 
interactions, they are highly conserved over the course of evolution.
Moreover, 
again because of these pairwise interactions, a change in one amino acid in a 
beta strand may require a change in a corresponding, hydrogen-bonded residue in 
another beta strand if the protein is to remain stable.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

These non-local interactions cause what had been a simple hidden Markov model 
to become a Markov random field.
SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
attempted to compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming.
Mathematically, we can represent the probability of a query sequence 
$x_{1}..x_{n}$ given a Markov random field with an additional term in the
Viterbi relations.
Specifically, each term $V_{j}^{\prime M}(i)$ becomes
$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log Pr(x_{i} | x_{\pi(j)})$, where
$x_{\pi(j)}$ is the amino acid in the query sequence that will be placed
in the beta-strand position paired with position $j$.
This nonlocal dependency prevents the problem from exhibiting optimal substructure,
preventing dynamic programming from solving the problem quickly.
SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand.
The algorithm underlying 
MRFy grew out of an understanding of these limitations.

>>>>

@@

@@

Standard implementations of the Viterbi algorithm, such as in HMMER, spend much of their code
managing the dynamic programming tables, as even when memoization libraries are available,
they are rarely used.
Furthermore, standard implementations in imperative languages typically avoid the natural
recursive representations because of the associated costs in languages such as C++.
The efficient machine code that GHC generates for recursive functions allowed us to write
a program that faithfully reflected the mathematics, without paying great costs in run-time
efficiency.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
simply duplicate 
functionality that is already implemented in the C++~package
HMMER.
These implementations form just one component of our new tool, MRFy.

@@

During the course of evolution, a mutation in an amino acid that participates
in this beta-strand pairing may cause a protein to fail to fold into a useful
structure, unless that mutation is accompanied by a \textit{compensatory}
mutation in the amino acid with which it is paired. 
Thus, homologous
proteins---those that share structure and function---tend to \textit{conserve}
such pairwise-conditional amino acids.

In order to model these non-local interactions, computational biologists
have extended hidden Markov models to include these non-local interactions,
modeled as a \textit{dependency graph}, which transforms the model into a
\textit{Markov random field}. These approaches ``nail down'' 
certain subsequences of the model, corresponding to the non-local hydrogen-bond 
interactions between beta strands.
The architecture of the 
resulting graphical model, derived from the HMMER ``Plan7'' architecture, is 
illustrated in Figure~\ref{mrf}.

SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming.
Mathematically, we can represent the probability of a query sequence 
$x_{1}..x_{n}$ given a Markov random field with an additional term in the
Viterbi relations.
Specifically, each term $V_{j}^{\prime M}(i)$ becomes
$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log Pr(x_{i} | x_{\pi(j)})$, where
$x_{\pi(j)}$ is the amino acid in the query sequence that will be placed
in the beta-strand position paired with position $j$.
This nonlocal dependency prevents the problem from exhibiting optimal substructure,
preventing dynamic programming from solving the problem quickly.
SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand.
The algorithm underlying 
MRFy grew out of an understanding of these limitations.
